\pagestyle{empty}
\selectlanguage{ngerman}
\noindent
\section*{Acknowledgement}
At this point, I would like to express my gratitude to all of those who directly or indirectly contributed to the successful completion of this thesis.

First and foremost, I would like to thank Dr.~Frank Scholze, head of the EUV radiometry group at the Physikalisch-Technische Bundesanstalt, for giving me the chance to conduct the work leading to this PhD thesis under his supervision. Our numerous scientific discussions, his valuable ideas and his constructive criticism bundled with the opportunity to conduct experiments even on a short notice, contributed significantly to the success of this thesis.

Furthermore, I would like to thank Prof.~Dr.~Mathias Richter for his support and the examination of this thesis. He always had an open ear and valuable advise for the course of my scientific work and the near future. 

I am very grateful to Prof.~Dr.~Stefan Eisebitt for supporting and evaluating my dissertation and to Dr.~Sa\v{s}a Bajt for the fruitful discussions and collaboration, for providing me with the Cr/Sc samples for my experiments and her willingness to serve as evaluator of this thesis. In addition, I thank Dr.~Stefan Braun for contributing the Mo/Si multilayer mirror samples.

I also like to acknowledge all of my current and former colleagues and fellow graduate students, first of all my mentor, Dr.~Victor Soltwisch, who supported me during the past years. I would also like to thank Anal\'{i}a Fern\'{a}ndez Herrero, Raül Garc\'{i}a Diez, Dr.~Christian Gollwitzer, Dr.~Philipp Hönicke, Mika Pflüger and Dr.~Jan Wernecke. Our many intense discussions and the collaborative atmosphere they helped to establish improved my research significantly.

I am sincerely grateful to all members of the working group 7.12, Christian Buchholz, Ayhan Babalik, Anja Babuschkin, Martin Biel, Benjamin Dubrau, Andreas Fischer, Anne Hesse, Sina Jaroslawzew, Florian Knorr, Dr.~Christian Laubis, Jana Lehnert, Heiko Mentzel, Jana Puls, Anja Schönstedt, Christian Stadelhoff. Without their support and patience in many late-night beamtimes in the laboratory, this work would not have been possible. My honest thanks also go to all other colleagues of the PTB in Berlin-Adlershof.

Finally, I am in dept to all of my friends and family for their endless support and their distractions during my studies and over the course of my PhD thesis. Most importantly I would like to name Michl, Michael, Paul, Laura, Tim, Anna, Leo and my parents Detlev and Martina. Last but not least, I am deeply grateful to my grandfather Dr.~Walther Neudert for inspiring me and his early encouragement of my scientific career.








\cleardoublepage
