%% publication list
\noindent
\pagestyle{empty}
\selectlanguage{ngerman}

\section*{Declaration}

Parts of this dissertation were previously published in peer-reviewed journals and conference contributions. I attach information on previous
publications according to ยง2 (4) of the \emph{Promotionsordnung} of TU Berlin and regulations of
Faculty II.

\vspace{2ex}

List of publications containing parts of the dissertation and
the detailed contributions of the co-authors to each publication:

\begin{enumerate}[label=\arabic*) ]
    \item \fullcite{haase_role_2014} 
	
	VS and FS concieved the study. AH conducted the EUV reflectivity and diffuse scattering measurements, developed the software, performed the data analysis and wrote the manuscript. VS, CL and FS contributed trough discussion of the results.
	All authors read, approved and contributed to the final manuscript.

    \item \fullcite{haase_characterization_2015} 
    
	AH, FS and SB designed the study. FS organized the samples. SB fabricated the samples and conducted the XRR measurements. AH conducted the EUV reflectivity and diffuse scattering measurements, analyzed the data and drafted the manuscript. VS and FS contributed trough discussion of the results. All authors read, approved and contributed to the final manuscript.

    \item \fullcite{haase_multiparameter_2016} 
    
	AH, SB and FS developed the study. AH organized the samples, conducted the EUV reflectivity, resonant EUV reflectivity and diffuse scattering measurements. SB fabricated the samples and conducted the XRR measurement. PH and AH performed the XRF experiment. AH developed the model and the corresponding software, analyzed the data of all experiments and wrote the manuscript. VS and FS contributed through discussion of the results. All authors read, approved and contributed to the final manuscript.

    \item \fullcite{haase_interface_2017} 
    
	AH, FS and SB designed the study. FS organized the samples. SB fabricated the samples and conducted the XRR measurements. AH conducted the EUV reflectivity and diffuse scattering measurements, analyzed the data, developed the software and drafted the manuscript. VS, CL and FS contributed trough discussion of the results. All authors read, approved and contributed to the final manuscript.

\end{enumerate}
List of publications not part of the dissertation:
\begin{enumerate}[label=\arabic*) ]
    \item \fullcite{prasciolu_extended_2015} 

    \item \fullcite{soltwisch_correlated_2016} 
    
\end{enumerate}

\vspace{3cm}

\noindent Berlin, den \today \hfill Anton Haase

\cleardoublepage
